---
title: "Problem Set #1 STATS 506"
author: "Jun Su"
format:
  html:
    embed-resources: true
    code-fold: true
---

### Q1

## a.Import the data

```{r}

soy <- read.csv("data/data.csv")

 
names(soy) <- c("Season", "Cultivar", "Repetition",
                "PH", "IFP", "NLP", "NGP",
                "NGL", "NS", "MHG", "GY") #Change the names accordingly

head(soy)
```
The dataset was imported from data.csv, and the variables were named according to the dataset documentation.

## b. verification

```{r}
#Number of observations
num_of_obs<-dim(soy)[1]
num_of_obs
#Number of distinct cultivars
length(unique(soy$Cultivar))

#Distinct seasons
unique(soy$Season)

#Number of observations in each season
table(soy$Season)
```
The dataset contains 320 observations, which represent 40 distinct soybean cultivars. The observations were collected during two unique seasons. Both have 160 observations

## c.(1) Find the correlation between PH and GY
```{r}
cor_PH_GY <- cor(soy$PH, soy$GY)
cor_PH_GY
```
The correlation is between plant height and grain yield 0.1232807, a positive association but not very high.

## c.(2) Find the correlation between PH and GY for each unique season.
```{r}
cor_season1 <- cor(
  soy$PH[soy$Season == 1],
  soy$GY[soy$Season == 1]
)

cor_season2 <- cor(
  soy$PH[soy$Season == 2],
  soy$GY[soy$Season == 2]
)

cor_season1
cor_season2
```
The correlation for season 1 yield 0.2308581, a positive association. However, the correlation for season 2 yield -0.0227046, which indicates a very weak negative correlation. Season 1 has a higher correlation.

## c. (3) The grain yield of the soybean observation with the greatest plant height
```{r}
soy_high<- soy[order(soy$PH, decreasing = TRUE), ]
soy_high[1,]$GY
```
The grain yield of the soybean observation with the greatest plant height is 3304.678

## c. (4) Percentage of observations that have a thousand seed weight (MHG) greater than the overall average thousand seed weight
```{r}
avg_MHG <- mean(soy$MHG)
avg_MHG
heavy_soy <- soy[soy$MHG > avg_MHG,]
perc_larger_weight <- dim(heavy_soy)[1]/num_of_obs *100
perc_larger_weight
```
The percentage of observations that have a thousand seed weight (MHG) greater than the overall average thousand seed weight is 47.1875%.

## c. (5) cultivar with the highest average grain yield
```{r}
avg_GY_by_cultivar <- aggregate(GY ~ Cultivar,
                                data = soy,
                                FUN = mean)
highest_GY_cultivar <- avg_GY_by_cultivar[order(avg_GY_by_cultivar$GY,decreasing = TRUE),][1,]
highest_GY_cultivar
```
The cultivar with the highest average grain yield is 98R30 CE with an average GY of 3964.968.

## d. create a table
```{r}
quant_vars <- c(
  "PH", "IFP", "NLP", "NGP",
  "NGL", "NS", "MHG", "GY"
)
quant_vars
```
These are quantitative variables: "PH"  "IFP" "NLP" "NGP" "NGL" "NS"  "MHG" "GY" 
```{r}
overall_avg <- colMeans(soy[quant_vars])

overall_avg
season1_avg <- colMeans(
  soy[soy$Season == 1, quant_vars]
)

season1_avg
season2_avg <- colMeans(
  soy[soy$Season == 2, quant_vars]
)

season2_avg
final_table <- rbind(overall_avg,season1_avg,season2_avg)
final_table
```

## e. t-test
First, define my own t test function
```{r}
#' Compute a two-sample t-test manually
#'
#' @param x,y 2 arrays containing the observations from the two groups.
#' @return A named numeric vector containing the sample means of x and y, t statistic, degrees of freedom,
#'   and p-value.
my_t_test <- function(x, y) {
  
  # sample sizes
  n1 <- length(x)
  n2 <- length(y)
  
  # sample means
  mean1 <- mean(x)
  mean2 <- mean(y)
  
  # sample var 
  var1 <- var(x)
  var2 <- var(y)
  
  # standard error
  SE <- sqrt(var1 / n1 + var2 / n2)
  
  # t statistic
  t <- (mean1 - mean2) / SE
  
  # dof
  df <- (var1 / n1 + var2 / n2)^2 /
        ((var1 / n1)^2 / (n1 - 1) +
         (var2 / n2)^2 / (n2 - 1))
  
  # p value
  pv <- 2 * pt(-abs(t), df = df) # look up the table and gives the 2 sided p value
  
  # results
  return(c(
    mean_1 = mean1,
    mean_2 = mean2,
    t_statistic = t,
    df = df,
    p_value = pv
  ))
}
```

```{r}
GY_1 <- soy$GY[soy$Season == 1]
GY_2 <- soy$GY[soy$Season == 2]
my_t_test(GY_1, GY_2)
```


We test
\(H_0: \mu_1 = \mu_2\), meaning that the mean grain yield is equal between the two seasons,
against
\(H_A: \mu_1 \neq \mu_2\), meaning that the mean grain yield differs between the two seasons.
Our p value is  0.725498>0.05, fail to reject H0. There is insufficient evidence that the mean grain yield differs between the 2 seasons.

### Q2

## a & b. load the data and clean the names


```{r}
adult <- read.csv(
  "data/adult.data",
  header = FALSE,
  strip.white = TRUE,
  na.strings = "?"
 )

names(adult) <- c(
  "age",
  "workclass",
  "final_weight",
  "education_level",
  "education_years",
  "marital_status",
  "occupation",
  "relationship",
  "race",
  "sex",
  "capital_gain",
  "capital_loss",
  "hours_per_week",
  "native_country",
  "income"
)
 
head(adult)
```
The Adult Income data were imported without a header, with surrounding whitespace removed and `?` treated as missing. The 15 columns were assigned the names shown above.

## c 
```{r}
for(name in names(adult)){
print(c(name,sum(is.na(adult[[name]]))))
}
```
There are 0 missing values in "age", 1836 missing values in "workclass", 0 missing values in "final_weight", 0 missing values in "education_level", 0 missing values in "education_years", 0 missing values in "marital_status", 1843 missing values in "occupation", 0 missing values in "relationship", 0 missing values in "race", 0 missing values in "sex", 0 missing values in "capital_gain", 0 missing values in "capital_loss", 0 missing values in "hours_per_week", 583 missing values in "native_country", and 0 missing values in "income".

## d

```{r}
sample_size_ori <- dim(adult)[1]
sample_size_ori
adult_long_work <- adult[adult$hours_per_week >= 10,]
sample_size_long <- dim(adult_long_work)[1]
sample_size_long

```
The original sample size is 32561, while the sample size of individuals working at least 10 hours per week is 32103.

## e

```{r}
adult_na_removed <- adult_long_work[
  !is.na(adult_long_work$workclass) &
  !is.na(adult_long_work$occupation) &
  !is.na(adult_long_work$native_country),
]
colSums(is.na(adult_na_removed[c("occupation", "workclass", "native_country")]))
```
Observations with missing values in occupation, workclass, or native_country were removed because these observations do not provide usable information for analyses involving those variables.

## f

```{r}
real_adult <- adult_na_removed[
  adult_na_removed$age >= 18 &
  adult_na_removed$age <= 90 &
  adult_na_removed$education_level != "Preschool",
]
range(real_adult$age)
sum(real_adult$education_level == "Preschool")
```
The retained observations have ages from 18 through 90, inclusive, and exclude the Preschool education category.

## g

```{r}
final_sample_size <- dim(real_adult)[1]
final_sample_size
```
The final sample size is 29523.

### Q3

```{r}
#' Compute the next value in a happy-number sequence
#'
#' @param n A single positive integer.
#' @return A positive integer equal to the sum of the squared digits of n.
nextHappy <- function(n) {
  if (
  length(n) != 1 ||
  !is.numeric(n) ||
  is.na(n) ||
  !is.finite(n) ||
  n <= 0 ||
  n %% 1 != 0
) {
  stop("n must be a single positive integer.")
}
  x <- as.character(n)
  digits <- as.numeric(strsplit(x, "")[[1]])
  sum(digits^2)
}
#' Compute the full happy-number sequence
#'
#' @param n A single positive integer.
#' @return the sequence array and whether this is a happy number
happySequence <- function(n) {
  if (
  length(n) != 1 ||
  !is.numeric(n) ||
  is.na(n) ||
  !is.finite(n) ||
  n <= 0 ||
  n %% 1 != 0
) {
  stop("n must be a single positive integer.")
}
  isHappy <- TRUE
  res <- c(n)
  
  while (n != 1) {
    n <- nextHappy(n)
    
    if (n %in% res) {
      res <- c(res, n)
      isHappy <- FALSE
      break
    }
    
    res <- c(res, n)
  }
  
  return(list(
    sequence = res,
    isHappy = isHappy
  ))
}
results <- lapply(1:1000, happySequence)

isHappy <- sapply(results, function(x) x$isHappy)
seqLen <- sapply(results, function(x) length(x$sequence))
# 1. number of happy
numHappy <- sum(isHappy)

# 2. percentage of happy
perHappy <- numHappy/1000 * 100

# 3. starting value with longest sequence
maxLen <- max(seqLen[isHappy])
maxIndex <- which(isHappy & seqLen == maxLen)

#Test cases required:
nextHappy(19)
nextHappy(82)
happySequence(19)$sequence
happySequence(19)$isHappy
happySequence(4)$isHappy
#c
numHappy
perHappy
maxIndex
maxLen
```
All test cases passed. Among 1 to 1000, there're 143 happy numbers, which is 14.3%.
The maximum happy-sequence length is 7. Multiple starting values are tied for the longest happy sequence. They are: 356 365 379 397 469 496 536 563 635 649 653 694 739 793 899 937 946 964 973 989 998.

## Attribution of Sources

OpenAI ChatGPT was used to review my draft against the assignment
requirements and to explain the help documents. I wrote, reviewed, and verified the final code
myself.
